---
title: "EDA"
output: pdf_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r packages}
# install.packages("f1dataR")
# install.packages("dplyr")
```

```{r setup-eda, message=FALSE, warning=FALSE}
library(f1dataR)
library(dplyr)
library(tidyverse)
```

```{r load-data}
# Load the pre-compiled dataset containing all seasons (2019–2024)
# full_data <- map_dfr(2019:2024, function(year) {
#   schedule <- load_schedule(season = year)
#   map_dfr(seq_len(nrow(schedule)), function(round) {
#     tryCatch(
#       load_results(season = year, round = round) %>%
#         mutate(
#           season = year,
#           round = round,
#           race_name = schedule$race_name[round],
#           circuit = schedule$circuit_id[round],
#           top_speed_kph = as.numeric(top_speed_kph),
#           fastest_rank = as.integer(fastest_rank)
#         ),
#       error = function(e) {
#         message("Skipping season ", year, " round ", round, ": ", e$message)
#         NULL
#       }
#     )
#   })
# })

# write_csv(full_data, "full_data.csv")

full_data <- read.csv("full_data.csv")

head(full_data, 8)
```
```{r helper-function}
# Helper function to convert gap values (strings like "+1:23.4") into numeric seconds
convert_to_seconds <- function(x) {
  if (is.numeric(x)) return(x)

  x <- as.character(x)

  sapply(x, function(val) {
    if (grepl("^[+-]?\\d+\\.?\\d*$", val)) {
      return(as.numeric(val))
    }

    if (!grepl("^[+-]", val) && grepl(":", val)) {
      return(0)
    }

    sign <- ifelse(startsWith(val, "-"), -1, 1)
    val <- gsub("^[+-]", "", val)

    parts <- strsplit(val, ":")[[1]]

    if (length(parts) == 2) {
      return(sign * (as.numeric(parts[1]) * 60 + as.numeric(parts[2])))
    } else if (length(parts) == 3) {
      return(sign * (as.numeric(parts[1]) * 3600 +
                     as.numeric(parts[2]) * 60 +
                     as.numeric(parts[3])))
    } else {
      return(NA_real_)
    }
  }, USE.NAMES = FALSE)
}
```

```{r clean-data}
# Clean dataset: convert gap to seconds, fix grid type, and filter valid race finishes
full_data <- full_data %>%
  mutate(
    gap_seconds = convert_to_seconds(gap),
    grid = as.integer(grid)
  ) %>%
  filter(
    status == "Finished",
    !is.na(grid),
    grid > 0,
    !is.na(gap_seconds)
  )

head(full_data, 8)
```

```{r plot}
# grid vs gap across all circuits
full_data %>%
  ggplot(aes(x = factor(grid), y = gap_seconds)) +
  geom_boxplot(outlier.alpha = 0.3, fill = "#2f2f2f", color = "#e10600") +
  scale_y_continuous(labels = scales::label_number(suffix = "s")) +
  labs(
    title = "Gap to Race Winner by Grid Position (2019–2024)",
    x = "Grid Position",
    y = "Gap to Winner (seconds)"
  ) +
  theme_minimal()
```
Gap to winner increases monotonically with grid position. Drivers starting 
from the front consistently finish closer to the winner, while higher grid 
positions show both larger median gaps and greater spread.

```{r}
# grid vs gap across all circuits scatterplot
full_data %>%
  ggplot(aes(x = grid, y = gap_seconds)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_x_continuous(breaks = 1:20) +
  scale_y_continuous(labels = scales::label_number(suffix = "s")) +
  scale_color_viridis_d(option = "turbo") +  # 20+ distinct colors
  labs(
    title = "Grid Position vs Gap to Winner (2019–2024)",
    x = "Grid Position",
    y = "Gap to Winner (seconds)"
  ) +
  theme_minimal()
```
The upward trend is clear but highly variable. Notable is the presence of 
gap = 0 observations across all grid positions, confirming that any starting 
position can produce a race winner.

```{r non-linearity, warning= FALSE}
# Nonlinearity check: overlays loess smoother vs linear fit to assess 
# whether grid position has a nonlinear effect on gap to winner

full_data %>%
  ggplot(aes(x = grid, y = gap_seconds)) +
  geom_point(alpha = 0.2, size = 0.8, color = "#e10600") +
  geom_smooth(method = "loess", color = "blue", 
              se = TRUE, linewidth = 1) +
  geom_smooth(method = "lm", color = "black", 
              linetype = "dashed", se = FALSE, linewidth = 1) +
  scale_x_continuous(breaks = 1:20) +
  labs(
    title = "Nonlinearity check: loess vs linear fit",
    subtitle = "Blue = loess smoother, dashed = linear fit",
    x = "Grid Position",
    y = "Gap to Winner (seconds)"
  ) +
  theme_minimal()
```

The loess smoother (blue) diverges from the linear fit (dashed black line), 
suggesting the relationship between grid position and gap to winner is not 
linear. The gap increases steeply from positions P1 to P8 but begins to 
flatten beyond P10, indicating diminishing returns at higher grid positions. 
This pattern suggests that a log transformation of the grid predictor may 
better capture the curvature in the data compared to a simple linear term.


```{r}
# average gap of each grid by circuit
full_data %>%
  group_by(circuit, grid) %>%
  summarise(avg_gap = mean(gap_seconds), .groups = "drop") %>%
  ggplot(aes(x = grid, y = circuit, fill = avg_gap)) +
  geom_tile(color = "#1a1a2e", linewidth = 0.3) +
  scale_x_continuous(breaks = 1:20) +
  scale_fill_gradientn(
    colors = c("#00d2ff", "#ffffff", "#e10600"),
    values = scales::rescale(c(0, 30, 120)),  # adjust midpoint if needed
    labels = scales::label_number(suffix = "s"),
    name = "Avg Gap"
  ) +
  labs(
    title = "Average Gap to Winner by Grid Position & Circuit (2019–2024)",
    x = "Grid Position",
    y = NULL
  ) +
  theme_minimal()
```

Average gap varies across circuits, suggesting that circuit characteristics play a role in determining finishing gaps. For example, high-speed circuits like Spa and Silverstone tend to produce larger gaps, while street circuits like Monaco and Interlagos appear more compressed. This variability across circuits suggests that a single pooled model may be insufficient, motivating the use of a hierarchical model with circuit-specific effects.

```{r, fig.height= 6}
# Sample size per circuit
full_data %>%
  count(circuit) %>%
  ggplot(aes(x = reorder(circuit, n), y = n)) +
  geom_col(fill = "#a93226") +
  coord_flip() +
  labs(
    title = "Number of observations per circuit (2019–2024)",
    x = NULL,
    y = "Number of observations"
  ) +
  theme_minimal()
```

The number of observations varies considerably across circuits. Established 
circuits such as Silverstone, Spa, and Bahrain appear across all six seasons 
and therefore have the most data, while newer or one-off circuits such as 
Mugello, Hockenheimring, and Nurburgring appear only once or twice. This 
imbalance means that estimates for data-sparse circuits would be unreliable 
if modelled independently. A hierarchical model addresses this by allowing 
circuits with fewer observations to borrow strength from the overall 
distribution, producing more stable estimates.

```{r fig.height= 6}
# grid vs gap for each circuit
full_data %>%
  ggplot(aes(x = grid, y = gap_seconds)) +
  geom_point(alpha = 0.6, size = 0.8, color = "#e10600") +
  facet_wrap(~ circuit, ncol = 4) +
  scale_x_continuous(breaks = c(1, 10, 20)) +
  scale_y_continuous(labels = scales::label_number(suffix = "s")) +
  labs(
    title = "Grid Position vs Gap to Winner by Circuit (2019–2024)",
    x = "Grid Position",
    y = "Gap to Winner (seconds)"
  ) +
  theme_minimal() 

```

Looking at individual circuits, the strength of the grid-gap relationship 
appears to vary. Some circuits show a steeper positive trend while others 
appear relatively flat, which is consistent with the patterns seen in the 
heatmap. This further supports the inclusion of circuit-specific slope 
parameters in the hierarchical model.

```{r plots, fig.height= 7}
# grid vs gap by season
full_data %>%
  ggplot(aes(x = grid, y = gap_seconds)) +
  geom_point(alpha = 0.6, size = 0.8, color = "#e10600") +
  facet_wrap(~ season, ncol = 4) +
  scale_x_continuous(breaks = c(1, 10, 20)) +
  scale_y_continuous(labels = scales::label_number(suffix = "s")) +
  labs(
    title = "Grid Position vs Gap to Winner by Season",
    x = "Grid Position",
    y = "Gap to Winner (seconds)"
  ) +
  theme_minimal()
```

The grid-gap relationship appears broadly consistent across the 2019 to 2024 
seasons, with no obvious systematic changes over time. This suggests that 
pooling data across seasons is a reasonable assumption for our analysis.

```{r}
# grid vs gap by round
full_data %>%
  ggplot(aes(x = grid, y = gap_seconds)) +
  geom_point(alpha = 0.6, size = 0.8, color = "#e10600") +
  facet_wrap(~ round, ncol = 4) +
  scale_x_continuous(breaks = c(1, 10, 20)) +
  scale_y_continuous(labels = scales::label_number(suffix = "s")) +
  labs(
    title = "Grid Position vs Gap to Winner by Round",
    x = "Grid Position",
    y = "Gap to Winner (seconds)"
  ) +
  theme_minimal() 
```
No clear pattern emerges when examining gap by round number within a season. 
The variation observed across rounds likely reflects race-specific conditions 
such as weather or safety car periods rather than any structural effect of 
round order.

```{r}
 full_data %>%
  select(grid, gap_seconds) %>%
  summary()
```

Grid positions in the cleaned dataset range from 1 to 20 with a mean of 8.4, 
while gap to winner ranges from 0 to 148.8 seconds with a mean of 39.3 
seconds. The large difference between the mean and the maximum gap, along 
with the right-skewed histogram, suggests that a normal likelihood would 
be inappropriate for modelling this response variable.

```{r hist}
hist(full_data$gap_seconds,
     col = "#a93226",
     main = "Histogram of gap to winner",
     xlab = "Gap to Winner (seconds)")

hist(filter(full_data, grid == 4)$gap_seconds,
     col = "#a93226",
     main = "Histogram of gap to winner (grid = 4)",
     xlab = "Gap to Winner (seconds)")
```

The distribution of gap to winner is heavily right-skewed, with most 
observations concentrated below 50 seconds and a long tail extending 
toward 150 seconds. The response is also strictly non-negative, which 
rules out the use of a Normal likelihood and motivates the choice of 
a distribution with positive support such as the Exponential.

Even when conditioning on a single grid position, the gap distribution 
remains right-skewed and strictly positive. This suggests the skewness 
is not simply a consequence of mixing different grid positions and is 
a genuine feature of the data.

```{r qqplot}
par(mfrow = c(1, 2))

qqnorm(full_data$gap_seconds, 
       main = "Q-Q Plot of raw gap")
qqline(full_data$gap_seconds, col = "red")

qqnorm(log(full_data$gap_seconds + 1), 
       main = "Q-Q Plot of log(gap + 1)")
qqline(log(full_data$gap_seconds + 1), col = "red")

```

Both the raw gap and log-transformed gap deviate substantially from the 
normal reference line. The raw gap shows severe right skew, and while 
the log transformation reduces this, the departure from normality remains 
visible particularly in the lower tail. These results suggest that neither 
a Normal nor a log-Normal likelihood is well-suited to this data, further 
supporting the choice of an Exponential likelihood.

```{r expo, fig.width= 9}
#Exponential fit check by grid position and rate = 1/mean(gap)

full_data |>
  group_by(grid) |>
  mutate(lambda = 1 / mean(gap_seconds)) |>
  ungroup() |>
  ggplot(aes(x = gap_seconds)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "#2b2b2b",
                 alpha = 0.8,
                 colour    = "white",
                 linewidth = 0.3) +
  geom_line(aes(x = gap_seconds, y = dexp(gap_seconds, rate = lambda)),
            color = "#a93226",
            alpha = 0.6,
            linewidth = 0.8) +
  facet_wrap(~ grid, ncol = 4,
             labeller = label_bquote(grid == .(grid))) +
  labs(
    title = "Exponential fit check by grid position",
    subtitle = "Red curve = Exp(1/mean(gap)) overlay",
    x = "Gap to Winner (seconds)",
    y = "Density"
  ) +
  theme_minimal()
```
Using a lambda = 1/mean(gap) for each grid position 
separately, the Exponential distribution captures the overall shape of 
the data reasonably well, particularly at lower grid positions where the 
fit is closest. The fit is less precise at higher grid positions, which 
is expected given the mean-variance constraint of the Exponential 
distribution. Overall, this provides reasonable support for the 
Exponential likelihood as a working approximation.

```{r, fig.height= 6}
# Exponential overlay function to test different alpha/beta values
# To understand reasonable priors
plot_exp_overlay <- function(alpha, beta) {
  params <- tibble(grid = 1:20) |>
    mutate(rate = alpha + beta * grid)
  
  overlay <- params |>
    rowwise() |>
    mutate(
      x    = list(seq(0, quantile(full_data$gap_seconds[full_data$grid == grid], 0.995),
                      length.out = 300)),
      dens = list(dexp(unlist(x), rate = rate))
    ) |>
    unnest(c(x, dens))
  
  ggplot(full_data, aes(x = gap_seconds)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins      = 30,
      fill      = "#2b2b2b",
      alpha     = 0.8,
      colour    = "white",
      linewidth = 0.3
    ) +
    geom_line(
      data      = overlay,
      aes(x = x, y = dens),
      colour    = "#a93226",
      alpha     = 0.8,
      linewidth = 0.8
    ) +
    facet_wrap(~ grid, scales = "free", ncol = 4,
               labeller = label_bquote(grid == .(grid))) +
    labs(
      x        = "gap_seconds",
      y        = "density",
      title    = "Exponential fit check by grid position",
      subtitle = sprintf("Exp(rate = %.3f + %.4f × grid) overlay", alpha, beta)
    ) +
    theme_minimal(base_size = 11) +
    theme(
      strip.text       = element_text(size = 9, face = "bold"),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold")
    )
}

# Test 3 different ranges
plot_exp_overlay(alpha = 0.115, beta = -0.005)
plot_exp_overlay(alpha = 0.05,  beta = -0.001)
plot_exp_overlay(alpha = 0.25,  beta = -0.012)
plot_exp_overlay(alpha = 0.50,  beta = -0.020)

```
Three different values of alpha and beta are tested using a linear rate 
parameterization. In all cases the exponential curve fails to track the 
histogram well across all grid positions, as the linear predictor cannot 
adequately capture the concave relationship seen in the nonlinearity check. 
This reinforces the motivation for using a log(grid) transformation of 
the predictor.

```{r,fig.height= 6}
plot_exp_overlay <- function(alpha, beta) {
  params <- tibble(grid = 1:20) |>
    mutate(rate = exp(alpha + beta * log(grid)))
  
  overlay <- params |>
    rowwise() |>
    mutate(
      x    = list(seq(0, quantile(full_data$gap_seconds[full_data$grid == grid], 0.995),
                      length.out = 300)),
      dens = list(dexp(unlist(x), rate = rate))
    ) |>
    unnest(c(x, dens))
  
  ggplot(full_data, aes(x = gap_seconds)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins      = 30,
      fill      = "#2b2b2b",
      alpha     = 0.8,
      colour    = "white",
      linewidth = 0.3
    ) +
    geom_line(
      data      = overlay,
      aes(x = x, y = dens),
      colour    = "#a93226",
      alpha     = 0.8,
      linewidth = 0.8
    ) +
    facet_wrap(~ grid, scales = "free", ncol = 4,
               labeller = label_bquote(grid == .(grid))) +
    labs(
      x        = "gap_seconds",
      y        = "density",
      title    = "Exponential fit check by grid position",
      subtitle = sprintf("Exp(rate = exp(%.2f + %.2f × log(grid))) overlay", alpha, beta)
    ) +
    theme_minimal(base_size = 11) +
    theme(
      strip.text       = element_text(size = 9, face = "bold"),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold")
    )
}

# Test 3 different ranges
plot_exp_overlay(alpha = -2.5, beta = -0.5)
plot_exp_overlay(alpha = -2.0, beta = -0.3)  
plot_exp_overlay(alpha = -3.0, beta = -0.7)
```

Replacing the linear predictor with log(grid) produces a better visual 
fit across grid positions compared to the linear version. Among the 
values tested, alpha around -2.5 and beta around -0.5 appear to provide 
the most reasonable coverage of the observed data, and these values 
will inform our prior specification in the Stan model.