---
title: "Proposal"
subtitle: "Bayesian F1 Grid Gap Analysis" 
author: "Shahrukh Islam Prithibi (47610068), Khue Do"
date: "2026-03-06"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r packages}
# install.packages("f1dataR")
# install.packages("dplyr")
```

## Team

Shahrukh Islam Prithibi, Khue Do

## Github Repo Link:

<https://github.com/khuedo22/stat-405-final-project.git>

## Datasets

### Primary Dataset

We use the `f1dataR` R package which pulls data directly from the
Jolpica/Ergast F1 API. The dataset contains race results for every F1
season from 1950–2024. For our analysis we focus on the 2019–2024
seasons. Each row corresponds to one driver's result in one race.

URL: <https://github.com/SCasanova/f1dataR>

The key variables we use are:

-   `driver_id` — unique identifier for the driver
-   `constructor_id` — the team the driver races for\
-   `grid` — starting grid position (1–20)
-   `gap` — finishing time behind the race winner in seconds
-   `position` — finishing position
-   `status` — whether the driver finished or retired (used for
    filtering)
-   `circuit` — the circuit where the race took place (from schedule)
-   `race_name` — the name of the Grand Prix (from schedule)

```{r data, message=FALSE, warning=FALSE}
library(f1dataR)
library(dplyr)

# Load schedule to get circuit and race name
schedule <- load_schedule(season = 2023)

sample <- load_results(season = 2023, round = 1) |>
  mutate(
    season        = 2023,
    round         = 1,
    race_name     = schedule$race_name[1],
    circuit       = schedule$circuit_id[1],
    top_speed_kph = as.numeric(top_speed_kph)
  )

head(sample)

```

### Backup Dataset

Our backup is a Kaggle dataset containing F1 race results as
pre-downloaded CSV files, available via our GitHub repository.

URL:
<https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020>

The relevant files are `results.csv`, `races.csv`, `drivers.csv`, and
`constructors.csv` which join on `raceId`, `driverId`, and
`constructorId` respectively. Unlike the primary dataset, this source
does not contain a `gap` column directly. Instead, we engineer it from
the `milliseconds` column in `results.csv`, which records each driver's
total race time in milliseconds. We compute the gap by subtracting the
winner's race time from each driver's race time and converting to
seconds:

```         
gap = (driver_milliseconds - winner_milliseconds) / 1000
```

```{r, eval=FALSE, message=FALSE, warning=FALSE}
base <- "https://raw.githubusercontent.com/khuedo22/stat-405-final-project/main/data_backup/"

results      <- read.csv(paste0(base, "results.csv"))
races        <- read.csv(paste0(base, "races.csv"))
drivers      <- read.csv(paste0(base, "drivers.csv"))
constructors <- read.csv(paste0(base, "constructors.csv"))

f1_backup <- results |>
  left_join(races |> rename(race_name = name),       by = "raceId") |>
  left_join(drivers,                                  by = "driverId") |>
  left_join(constructors |> rename(team_name = name), by = "constructorId") |>
  select(season = year, round, raceId, driverId,
         driver = surname, team = team_name,
         race_name, grid, position, points,
         milliseconds) |>
  filter(season == 2023) |>
  mutate(milliseconds = as.numeric(milliseconds)) |>  # force to numeric
  group_by(raceId) |>
  mutate(
    winner_ms = min(milliseconds, na.rm = TRUE),
    gap = (milliseconds - winner_ms) / 1000
  ) |>
  ungroup()

head(f1_backup)

```

## Question of interest

Does grid position affect a driver's finishing gap to the race winner,
and does this effect vary by circuit?

Grid position in F1 is set during qualifying: the faster you go, the
further up the grid you start on race day. Starting further back
generally means more cars to get past and a higher chance of getting
stuck in traffic, which should translate to finishing further behind the
winner. But how much this actually matters is probably not the same
everywhere. At a place like Monaco, the streets are so narrow that
overtaking is nearly impossible, so where you start is pretty much where
you finish. Monza is the opposite: long straights and DRS zones mean
cars can pass each other more freely, so a driver starting 10th still
has a realistic shot at a top finish.

This leads to our question:

> *Does grid position affect a driver's finishing gap to the race
> winner, and does this effect vary by circuit?*

We want to estimate how much of a time penalty a driver faces for each
grid position they start behind the front and whether that penalty is
bigger at some circuits than others. We think Bayesian methods are a
natural fit here since each circuit only has one race per season, so we
are working with limited data per group. A hierarchical model lets
circuits with less data borrow information from the broader pool of
circuits rather than being estimated in isolation, which gives more
reliable estimates especially for newer circuits on the calendar.

## Methodology

The parameter of interest will be the coefficient for a driver's
starting position. We plan to fit a Bayesian generalized regression
model. Since our response variable (`position`) is a discrete value from
1-10, we might model this either as a Categorical distribution. The
function for the model is $\alpha \times \text{grid}$. $p_i$ will the
the logit of the function. The parameters of interest are the
coefficient for grid position. This could follow a normal distribution.
We can build a hierarchial model for the parameters of the normal
distribution for the coefficient.

## Plan

-   data preprocessing
-   EDA
-   define some plausible models (distributions of data and parameters)
-   run model in STAN
-   build credible intervals and check whether they are well-calibrated
    (by predicting the next data point)
-   model rank histogram and trace plots, other diagnostics
-   iterate if needed
