---
title: "Proposal"
author: "Shahrukh Islam Prithibi, Khue Do"
date: "2026-03-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Team
Shahrukh Islam Prithibi, Khue Do

## Datasets
### Primary Dataset
F1 dataset R package: https://github.com/SCasanova/f1dataR 

This R package contains several datasets on F1 from 1950-2024. Datasets are subset by season and round (race number). Some information included are the lap times for each driver and driver telmetry. Below is the dataset of interest (I think), where each row contains information about how a driver performed in a particular race. 

The columns in this table are:  TODO provide a description for these variables Shahrukh pls verify these descriptions :>

- `driver_id` The name of the driver that is racing
- `constructor_id` The car they are a part of
- `points` How many points they earned for this race, which is based on `position`. Drivers who rank below 10th place get no points.
- `position` Where the driver finished in the race (from 1-20)
- `grid` Where a driver starts in the race
- `laps`
- `status`
- `gap`
- `fastest_rank`
- `fastest`
- `top_speed_kph`
- `time_sec`

```{r}
library(f1dataR)

head(load_results(season="current", round="last"))
```

### Backup Dataset

tbd

## Question of interest
How does a driver's starting position (`grid`) influence where they finish in the race (`position`)? 

## Methodology
The parameter of interest will be the coefficient for a driver's starting position. We plan to fit a Bayesian generalized regression model. Since our response variable (`position`) is a discrete value from 1-10, we might model this either as a Categorical distribution. The function for the model is $\alpha \times \text{grid}$. $p_i$ will the the logit of the function. The parameters of interest are the coefficient for grid position. This could follow a normal distribution.  We can build a hierarchial model for the parameters of the normal distribution for the coefficient.


## Plan
tbd