---
title: "Proposal"
subtitle: "Bayesian F1 Grid Gap Analysis" 
author: "Shahrukh Islam Prithibi (47610068), Khue Do (70790423)"
date: "2026-03-06"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r packages}
# install.packages("f1dataR")
# install.packages("dplyr")
```

## Team

Shahrukh Islam Prithibi, Khue Do

## Github Repo Link:

<https://github.com/khuedo22/stat-405-final-project.git>

## Datasets

### Primary Dataset

We use the `f1dataR` R package which pulls data directly from the
Jolpica/Ergast F1 API. The dataset contains race results for every F1
season from 1950–2024. For our analysis we focus on the 2019–2024
seasons. Each row corresponds to one driver's result in one race.

URL: <https://github.com/SCasanova/f1dataR>

The key variables we use are:

-   `driver_id` — unique identifier for the driver
-   `constructor_id` — the team the driver races for\
-   `grid` — starting grid position (1–20)
-   `gap` — finishing time behind the race winner in seconds
-   `position` — finishing position
-   `status` — whether the driver finished or retired (used for
    filtering)
-   `circuit` — the circuit where the race took place (from schedule)
-   `race_name` — the name of the Grand Prix (from schedule)

```{r data, message=FALSE, warning=FALSE}
library(f1dataR)
library(dplyr)

# Load schedule to get circuit and race name
schedule <- load_schedule(season = 2023)

sample <- load_results(season = 2023, round = 1) |>
  mutate(
    season        = 2023,
    round         = 1,
    race_name     = schedule$race_name[1],
    circuit       = schedule$circuit_id[1],
    top_speed_kph = as.numeric(top_speed_kph)
  )

head(sample)

```

### Backup Dataset

Our backup is a Kaggle dataset containing F1 race results as
pre-downloaded CSV files, available via our GitHub repository.

URL:
<https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020>

The relevant files are `results.csv`, `races.csv`, `drivers.csv`, and
`constructors.csv` which join on `raceId`, `driverId`, and
`constructorId` respectively. Unlike the primary dataset, this source
does not contain a `gap` column directly. Instead, we engineer it from
the `milliseconds` column in `results.csv`, which records each driver's
total race time in milliseconds. We compute the gap by subtracting the
winner's race time from each driver's race time and converting to
seconds:

```         
gap = (driver_milliseconds - winner_milliseconds) / 1000
```

```{r, eval=FALSE, message=FALSE, warning=FALSE}
base <- "https://raw.githubusercontent.com/khuedo22/stat-405-final-project/main/data_backup/"

results      <- read.csv(paste0(base, "results.csv"))
races        <- read.csv(paste0(base, "races.csv"))
drivers      <- read.csv(paste0(base, "drivers.csv"))
constructors <- read.csv(paste0(base, "constructors.csv"))

f1_backup <- results |>
  left_join(races |> rename(race_name = name),       by = "raceId") |>
  left_join(drivers,                                  by = "driverId") |>
  left_join(constructors |> rename(team_name = name), by = "constructorId") |>
  select(season = year, round, raceId, driverId,
         driver = surname, team = team_name,
         race_name, grid, position, points,
         milliseconds) |>
  filter(season == 2023) |>
  mutate(milliseconds = as.numeric(milliseconds)) |>  # force to numeric
  group_by(raceId) |>
  mutate(
    winner_ms = min(milliseconds, na.rm = TRUE),
    gap = (milliseconds - winner_ms) / 1000
  ) |>
  ungroup()

head(f1_backup)

```

## Question of interest

Does grid position affect a driver's finishing gap to the race winner,
and does this effect vary by circuit?

Grid position in F1 is set during qualifying: the faster you go, the
further up the grid you start on race day. Starting further back
generally means more cars to get past and a higher chance of getting
stuck in traffic, which should translate to finishing further behind the
winner. But how much this actually matters is probably not the same
everywhere. At a place like Monaco, the streets are so narrow that
overtaking is nearly impossible, so where you start is pretty much where
you finish. Monza is the opposite: long straights and DRS zones mean
cars can pass each other more freely, so a driver starting 10th still
has a realistic shot at a top finish.

This leads to our question:

> *Does grid position affect a driver's finishing gap to the race
> winner, and does this effect vary by circuit?*

We want to estimate how much of a time penalty a driver faces for each
grid position they start behind the front and whether that penalty is
bigger at some circuits than others. We think Bayesian methods are a
natural fit here since each circuit only has one race per season, so we
are working with limited data per group. A hierarchical model lets
circuits with less data borrow information from the broader pool of
circuits rather than being estimated in isolation, which gives more
reliable estimates especially for newer circuits on the calendar.

## Methodology

First, we will preprocess the data and perform some EDA on the data,
visualizing the relationship between grid positions and time gap for
each circuit to understand the relationship.

The next step is to define some plausible models to use. A naive
approach we may take is to assume all circuits have the same effect and
model just the relationship between the grid position and the time gap.
Since the time gap is strictly positive, we will model the log of the
time gap + 1 to account for first place finishes (time gap of 0).

$$
\begin{aligned}
\log(1+\text{gap}_i) \sim N(\alpha + \beta * \text{grid}_i, \sigma^2) \\
\alpha \sim N(0, 50) \\
\beta \sim N(0, 10) \\
\sigma \sim Exp(1)
\end{aligned}
$$

Another option we could take is to model $gap_i$ with an exponential
distribution with parameter $\alpha + \beta * \text{grid}_i$

$\beta$ represents the average number of seconds behind the winner for
each additional grid position. This would be the latent variable we are
interested in.

To account for the difference in circuit, we can model this relationship
for each 25 circuits. We can also model this in a hierarchical structure
to reduce the sensitivity of the prior.

$$
\begin{aligned}
\log(1+\text{gap}_{i}) \sim N(\alpha + \beta_j * \text{grid}_i, \sigma^2) \\
\alpha \sim N(0, 50) \\
\beta \sim N(\mu, \sigma) \\
\sigma \sim Exp(1) \\
\mu_\beta \sim N(0, 10) \\
\sigma_\beta \sim Exp(1)
\end{aligned}
$$

$j$ represents a circuit.

We plan to compute the posterior of this model using MCMC in Stan and
build credible intervals. After, we will perform diagnostics to evaluate
the model, such as examining credible interval calibration via
prediction, performing prior and posterior checks, and examining rank
plots and trace plots. The results of these steps will determine our
next steps after, such as debugging any issues, such as slow mixing or
re-evaluating our model. This will also allow us to determine any
limitations of our model. After building and checking our model, we will
interpret our final results to see how grid position relates to the time
gap and whether this effect varies by circuit.

## Tentative Plan

-   literature review - Shahrukh 
-   problem formulation - Shahrukh and Khue
-   data preprocessing - Shahrukh
-   EDA - Khue
-   Define two plausible models to test - Shahrukh and Khue (one each)
-   run model in STAN - Shahrukh
-   correctness check (software testing) - Shahrukh
-   credible interval check - Khue
-   diagnostics and other model evaluation - Khue
-   model limitations - Khue
-   interpretation - Shahrukh and Khue
